Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-18 modelling/modelling.tex
Big fix

\begin{align}
    \label{eq:model1}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_N+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\lambda_\Phi+\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=\frac{s_{\beta N}N}{1+i_{\beta\Phi}\Phi}-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}

It was assumed that there are negative feedbacks the cells and cytokines, i.e. overcrowding inhibits the increasing rates. As sources, macrophage produces tnf-$\alpha$ and neutrophil produces il-1$\beta$; however, macrophage was assumed to have a negative effect on the il-1$\beta$, term $1+i_{\beta\Phi}\Phi$ in Eqn. \ref{eq:model1} was used to represent this. Despite the self-driven increase of immune cells, il-1$\beta$ was assumed to promote the increase of both cells. Neutrophil was also assumed to be inhibited by the total immune cells presented in the lesion site, the term $\nu_{N\Phi}N\Phi$ was used to denote this.

@@

Model 2 and model 3 introduced one extra parameter $a$ and removed $\lambda_\Phi$, which is a coefficient in the exponentially decaying $\lambda_N$, determining the decay speed, with the unit $h^{-1}$.

\begin{align}
    \label{eq:model2}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=\frac{s_{\beta N}N}{1+i_{\beta\Phi}\Phi}-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}

\begin{align}
    \label{eq:model3}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=s_{\beta N}N-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}

Model 3 can be regarded as a simplification of model 2, as it can be treated as model 2 with parameter $i_{\beta\Phi}=0$. Among the proposed three models, model 1 is a naive one that was proposed at the very first time and used as an ODE `template' to build and test our parameter inference framework. As the implementation was successful, model 2 and 3 was proposed, as we were trying to calibrate some terms and find better models. After fitting, model 2 is supposed to be better than model 1 as it corrects the problem that appears at the final time points (which were related to the steady states of immune cells). Model 3 made a small simplification based on model 2, and thus it was theoretically less `general' than model 2.

@@

After the first model selection experiment among the exiting three models, it was found that any of the models did not recover some significant features presented in the observed data. To resolve this, attempts were tried to introduce additional reasonable interactions within the dynamic system, considering the biological and mathematical context. Extra promoting effect to the expression of tnf-$\alpha$ was considered, by either adding a phenomenological term $d_{\beta\alpha}\beta$ (which means the same effect as directly promoting but the underlying mechanism is unclear), or adding a term $f_{\beta\alpha}$ that represents a promoting effect to the production process of tnf-$\alpha$, namely model 4 (Eqn. \ref{eq:model4}) and model 5 (Eqn. \ref{eq:model5}).

\begin{align}
    \label{eq:model4}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=s_{\beta N}N-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha+d_{\beta\alpha}\beta
    \end{split}
\end{align}

\begin{align}
    \label{eq:model5}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=s_{\beta N}N-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=(s_{\alpha\Phi}+f_{\beta\alpha}\beta)\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}

\begin{table}[h!]
    \centering
    \begin{tabular}{|c c c|}
        \hline
        Parameter         & Definition                                                     & Units                    \\ [0.5ex]
        \hline\hline
        $d_{\beta\alpha}$ & Coefficient of promoting effect from $\beta$                   & $h^{-1}$                 \\
        \hline
        $f_{\beta\alpha}$ & Coefficient of promoting the production of $\alpha$ by $\beta$ & $cell^{-1}\cdotp h^{-1}$ \\
        \hline
    \end{tabular}
    \caption{Parameters introduced in model 4 and 5}
    \label{table:m45}
\end{table}

>>>>

\end{figure}
\begin{align}
    \label{eq:model1}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_N+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\lambda_\Phi+\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=\frac{s_{\beta N}N}{1+i_{\beta\Phi}\Phi}-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}
It was assumed that there are negative feedbacks the cells and cytokines, i.e. overcrowding inhibits the increasing rates. As sources, macrophage produces tnf-$\alpha$ and neutrophil produces il-1$\beta$; however, macrophage was assumed to have a negative effect on the il-1$\beta$, term $1+i_{\beta\Phi}\Phi$ in Eqn. \ref{eq:model1} was used to represent this. Despite the self-driven increase of immune cells, il-1$\beta$ was assumed to promote the increase of both cells. Neutrophil was also assumed to be inhibited by the total immune cells presented in the lesion site, the term $\nu_{N\Phi}N\Phi$ was used to denote this.

@@

Model 2 and model 3 introduced one extra parameter $a$ and removed $\lambda_\Phi$, which is a coefficient in the exponentially decaying $\lambda_N$, determining the decay speed, with the unit $h^{-1}$.
\begin{align}
    \label{eq:model2}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=\frac{s_{\beta N}N}{1+i_{\beta\Phi}\Phi}-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}

\begin{align}
    \label{eq:model3}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=s_{\beta N}N-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}
Model 3 can be regarded as a simplification of model 2, as it can be treated as model 2 with parameter $i_{\beta\Phi}=0$. Among the proposed three models, model 1 is a naive one that was proposed at the very first time and used as an ODE `template' to build and test our parameter inference framework. As the implementation was successful, model 2 and 3 was proposed, as we were trying to calibrate some terms and find better models. After fitting, model 2 is supposed to be better than model 1 as it corrects the problem that appears at the final time points (which were related to the steady states of immune cells). Model 3 made a small simplification based on model 2, and thus it was theoretically less `general' than model 2.

@@

After the first model selection experiment among the exiting three models, it was found that any of the models did not recover some significant features presented in the observed data. To resolve this, attempts were tried to introduce additional reasonable interactions within the dynamic system, considering the biological and mathematical context. Extra promoting effect to the expression of tnf-$\alpha$ was considered, by either adding a phenomenological term $d_{\beta\alpha}\beta$ (which means the same effect as directly promoting but the underlying mechanism is unclear), or adding a term $f_{\beta\alpha}$ that represents a promoting effect to the production process of tnf-$\alpha$, namely model 4 (Eqn. \ref{eq:model4}) and model 5 (Eqn. \ref{eq:model5}).
\begin{align}
    \label{eq:model4}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=s_{\beta N}N-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha+d_{\beta\alpha}\beta
    \end{split}
\end{align}

\begin{align}
    \label{eq:model5}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=s_{\beta N}N-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=(s_{\alpha\Phi}+f_{\beta\alpha}\beta)\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}
\begin{table}[h!]
    \centering
    \begin{tabular}{|c c c|}
        \hline
        Parameter         & Definition                                                     & Units                    \\ [0.5ex]
        \hline\hline
        $d_{\beta\alpha}$ & Coefficient of promoting effect from $\beta$                   & $h^{-1}$                 \\
        \hline
        $f_{\beta\alpha}$ & Coefficient of promoting the production of $\alpha$ by $\beta$ & $cell^{-1}\cdotp h^{-1}$ \\
        \hline
    \end{tabular}
    \caption{Parameters introduced in model 4 and 5}
    \label{table:m45}
\end{table}

### 2020-08-18 performance/performance.tex
Fig fix
Big fix

    \caption[Speedup of the program]{Speedup of the program}
    \label{fig:speedup}
\end{figure}

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|c c c c c|}
        \hline
        Number of cores & Time (second) & Avg. required samples & Speedup & Efficiency \\ [0.5ex]
        \hline\hline
        8               & 6617          & 445406.8              & 1       & 1          \\
        16              & 3926          & 590221.0              & 1.61    & 0.807      \\
        24              & 2921          & 573339.8              & 2.17    & 0.723      \\
        36              & 2096          & 446948.7              & 3.02    & 0.672      \\
        54              & 2076          & 640404.9              & 3.05    & 0.452      \\
        72              & 2071          & 768096.0              & 3.06    & 0.340      \\
        \hline
    \end{tabular}
    \caption{Performance data}
    \label{table:performance}
\end{table}

@@

\begin{figure}[h]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/sample_per_time.png}}
    \end{center}

    \caption[Average samples in a second, under different number of cores]{Average samples in one second, under different number of cores}
    \label{fig:sample_per_sec}
\end{figure}

@@

Due to the variant required samples, our interests switched to the per-second performance, where the average sampling numbers per second under different of numbers of cores ($p$) are plotted (Figure \ref{fig:sample_per_sec}). The plot shows that even theoretical speedup stops growing when $p>36$, the average sampling numbers per second performance still get improved when using more than 36 cores. It proves that the implementation used here scales reasonably well without encountering bottleneck when using multi-core parallelisation inside one physical node.

@@

\begin{figure}[ht]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/local_modes.png}}
    \end{center}

    \caption[Local optimum and abnormal pattern]{(a) Required sampling size for two example runs. Different colour represents different generation (from bottom to top: generation 1 to generation 20). Run A took huge amount of time in population 11. (b) Required sample in each generation of Run A. (c) Time-consumption in each generation of Run A}
    \label{fig:local_modes}
\end{figure}

@@

\begin{figure}[t]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/local_para.png}}
    \end{center}

@@

Moreover, we found that the execution time was not ideally proportional to the sampling tires in that period of time, for both different repeated runs and different generations in a single run. From the same example, Figure \ref{fig:local_modes} (b) and (c) show the non-linear relationship between execution time and sample tries in different generations. Apart from 1st and 2nd generation, generation 3 to 20 show a general linear relationship i.e. $T_t\propto n_t$, where $t$ is the population index, $T_t$ is the execution time of generation $t$ and $n_t$ is the required samples in generation $t$. However, this relationship does not hold for the first two generations. Generation 1 is to draw samples from prior, but it is done with high efficiency as only a small amount of time are taken; generation 2 takes an unusually large amount of the, but it does not require much more samples. The abnormal execution time of generation 1 and two are observed in all other runs, where for $p=24$ generation 1 usually takes c. 16s and generation 2 takes c. 1600s. This is believed to be related to the initialisation of the algorithm.

Additionally, by observing CPU frequency and load curve in-time while running the program, we found that the parallelised part is limited to the sampling, calculation and criterion test process. Some reduction calculations and preparation work for the next generation (e.g. calculation of the new epsilon values, normalisation of data, fit of multivariate kernel) are serial. The database I/O is also serial.

>>>>

    \caption[Speedup of the program]{Speedup of the program. Line in orange indicates the ideal linear speedup}
    \label{fig:speedup}
\end{figure}

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|c c c c c|}
        \hline
        Number of cores & Avg. time (second) & Avg. sample size & Speedup & Efficiency \\ [0.5ex]
        \hline\hline
        8               & 6617          & 445406.8              & 1       & 1          \\
        16              & 3926          & 590221.0              & 1.61    & 0.807      \\
        24              & 2921          & 573339.8              & 2.17    & 0.723      \\
        36              & 2096          & 446948.7              & 3.02    & 0.672      \\
        54              & 2076          & 640404.9              & 3.05    & 0.452      \\
        72              & 2071          & 768096.0              & 3.06    & 0.340      \\
        \hline
    \end{tabular}
    \caption{Performance data}
    \label{table:performance}
\end{table}

@@

% \begin{figure}[h]
%     \begin{center}
%         \resizebox{1.0\hsize}{!}{\includegraphics{fig/sample_per_time.png}}
%     \end{center}

%     \caption[Average samples in a second, under different number of cores]{Average samples in one second, under different number of cores}
%     \label{fig:sample_per_sec}
% \end{figure}

@@

Due to the variant total required samples drawn our attention to the per-second performance. Although the execution time was not improved by using 54 or 72 cores, the required number of samples were higher than that of $p=36$, which indicates slight improvements. However, the high variance made us hard to gave statistically significant conclusions on the per-second performance.

% our interests switched to the per-second performance, where the average sampling numbers per second under different of numbers of cores ($p$) are plotted (Figure \ref{fig:sample_per_sec}). The plot shows that even theoretical speedup stops growing when $p>36$, the average sampling numbers per second performance still get improved when using more than 36 cores. It proves that the implementation used here scales reasonably well without encountering bottleneck when using multi-core parallelisation inside one physical node.

@@

\begin{figure}[ht!]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/local_modes_new.png}}
    \end{center}

    \caption[Local optimum and abnormal pattern]{(a) Required number of samples for two example runs (A and B). Different colour represents different generation (from bottom to top: generation 1 to generation 20). Run A took huge amount of time in population 11. (b) Execution time for each generation of Run A. (c) Scatter plot of required number of samples against execution time for each generation. Generation 1 and two outliers, i.e. generation 11 and generation 2, are marked in different colours from other generations}
    \label{fig:local_modes}
\end{figure}

@@

\begin{figure}[ht!]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/local_para.png}}
    \end{center}

@@

Moreover, we found that the execution time was not ideally proportional to the sampling tires in that period of time, for both different repeated runs and different generations in a single run. From the same example, the sample size for each generation of Run A (Figure \ref{fig:local_modes} (a)) is not proportional to its execution time (Figure \ref{fig:local_modes} (b)), especially for the first two generations and generation 11. Figure \ref{fig:local_modes} (c) shows a scatter plot of the required samples and execution time for each generation, where two obvious outliers (generation 1 and generation 11) are observed.

% Figure \ref{fig:local_modes} (b) and (c) show the non-linear relationship between execution time and sample tries in different generations.

% Apart from 1st and 2nd generation, generation 3 to 20 show a general linear relationship i.e. $T_t\propto n_t$, where $t$ is the population index, $T_t$ is the execution time of generation $t$ and $n_t$ is the required samples in generation $t$.

% However, this relationship does not hold for the first two generations.

Generation 1 (red pont in Figure \ref{fig:local_modes} (c)) drew samples from prior, but it was done with high efficiency as only a small amount of time are taken; generation 2 (purple pont in Figure \ref{fig:local_modes} (c)) took an unusually large amount of the while it did not require much more samples. The abnormal execution time of generation 1 and 2 were observed in all other runs, where for $p=24$ generation 1 usually takes c. 16s and generation 2 takes c. 1600s. This is believed to be related to the initialisation of the algorithm. For local modes, e.g. generation 11 (green pont in Figure \ref{fig:local_modes} (c)), high efficiency were achieved.

By looking into the code, some reasons can be found. There exist some `cheap' particles. If at the first time points, a variable reached unacceptable high value (regarded as \verb|Inf| in \verb|Python|), the ODE solver cannot perform further integration but result in a data full of \verb|NaN|s. Such particles cost less time than an usual particle (where values at 120 data points are explicitly calculated and then compared with observed data). Local mode, e.g. generation 11, might take large amount of these `cheap' particles. Also, generation 1 draws samples directly from prior without fitting and using kernels, so it is much faster. The huge amount of time observed in generation 2 is believed to be a result of some initialisation steps in the ABC SMC algorithm.

To conclude, the performance experiments revealed reasonable scaling-up behaviour when the \verb|Python| processes did not exceed number of physical cores. Local modes lead to high variance in total required number of samples, and the relationship between required sample numbers and required execution time are not ideally linear; they all made further detailed analysis difficult.

Additionally, by observing CPU frequency and load curve in-time while running the program, we found that the parallelised part is limited to the sampling, calculation and criterion test process. Some reduction calculations and preparation work for the next generation (e.g. calculation of the new epsilon values, normalisation of data, fit of multivariate kernel) are serial. The database I/O is also serial.  It is noticed that the program took no advantage of graphics card, as \verb|multiprocessing| along can only exploit the resources inside CPU.
